---
title: "Reference 4"
subtitle: "Combining Plots with `patchwork`"
author: "Byeong-Hak Choe"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 6
    number-sections: true
    theme: cosmo
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
  fig-height: 8
---


```{r setup}
#| include: false
library(tidyverse)
library(patchwork)
library(hrbrthemes)
library(ggthemes)

theme_set(
  theme_ipsum() +
    theme(
      strip.background = element_rect(fill = "lightgray"),
      axis.title.x = element_text(
        angle = 0,
        size = rel(1.5),
        margin = margin(10, 0, 0, 0)
      ),
      axis.title.y = element_text(
        # angle = 0,
        size = rel(1.5),
        margin = margin(0, 10, 0, 0)
      ),
      axis.text.x = element_text(size = rel(1.25)),
      axis.text.y = element_text(size = rel(1.25)),
      strip.text = element_text(size = rel(1.3)),
      legend.position = "top",
      legend.text = element_text(size = rel(1.2)),
      legend.title = element_text(size = rel(1.2))
    )
)

scale_colour_discrete <- function(...) scale_colour_tableau(...)
scale_fill_discrete <- function(...) scale_fill_tableau(...)
```

---

# Introduction to `patchwork`

When building a data story, we often need to place multiple plots side by side, stack them vertically, or arrange them in a custom grid. Base R's `par(mfrow = ...)` can do this for base graphics, but it does not work with `ggplot2`.

**`patchwork`** is the modern solution. It extends `ggplot2` with a simple, expressive syntax for combining plots — using everyday operators like `+`, `|`, and `/`.

> "The goal of patchwork is to make it ridiculously simple to combine separate ggplots into the same graphic."
> — Thomas Lin Pedersen (package author)

## Installation and loading

```{r}
#| eval: false
install.packages("patchwork")
# remotes::install_github("thomasp85/patchwork")
```

```{r}
library(ggplot2)
library(patchwork)
```

---

# Building Blocks — Four `mtcars` Plots

We create four plots from `mtcars` that we will combine in different ways throughout this lecture. Think of these as the building blocks.

```{r}
p1 <- ggplot(mtcars) +
  geom_point(aes(mpg, disp), color = "steelblue") +
  labs(title = "Plot 1", x = "Miles per gallon", y = "Displacement")

p2 <- ggplot(mtcars) +
  geom_boxplot(aes(factor(gear), disp, fill = factor(gear)),
               show.legend = FALSE) +
  labs(title = "Plot 2", x = "Gears", y = "Displacement")

p3 <- ggplot(mtcars) +
  geom_point(aes(hp, wt, color = mpg)) +
  scale_color_viridis_c() +
  labs(title = "Plot 3", x = "Horsepower", y = "Weight")

p4 <- ggplot(mtcars) +
  geom_bar(aes(factor(gear), fill = factor(gear)),
           show.legend = FALSE) +
  facet_wrap(~cyl) +
  labs(title = "Plot 4", x = "Gears", y = "Count")
```

Display them individually to understand what each one shows before combining.

```{r}
p1
```

```{r}
p2
```

```{r}
p3
```

```{r}
p4
```

---

# Basic Use — the `+` Operator

The simplest way to combine two plots is `+`. Patchwork places them side by side in a single row.

```{r}
p1 + p2
```

::: {.callout-note}
The `+` operator is the same one used in `ggplot2` to add layers, but when applied *between two complete `ggplot` objects*, `patchwork` intercepts it and creates a combined layout.
:::

## Additions flow to the last plot

When you add a `ggplot2` element (such as `labs()`) after a `patchwork`, it is applied to the **last** plot in the composition:

```{r}
p1 + p2 + labs(subtitle = "This subtitle appears in the last plot (p2)")
```

This is consistent with `ggplot2`'s own `+` semantics — keep it in mind to avoid accidental modifications.

---

# Filling a Grid

Adding more than two plots fills a grid automatically. Patchwork tries to keep the grid as square as possible, filling in **row-first** order.

```{r}
p1 + p2 + p3 + p4
```

## Controlling rows and columns with `plot_layout()`

Use `plot_layout()` to override the default grid dimensions.

```{r}
p1 + p2 + p3 + p4 +
  plot_layout(nrow = 3, byrow = FALSE)
```

| Argument | Effect |
|----------|--------|
| `nrow` | Number of rows |
| `ncol` | Number of columns |
| `byrow` | Fill by row (`TRUE`) or column (`FALSE`) |
| `widths` | Relative widths of columns |
| `heights` | Relative heights of rows |

### Relative sizing

Control how much space each column or row gets:

```{r}
p1 + p2 + p3 +
  plot_layout(ncol = 3, widths = c(2, 1, 1))
```

Here `p1` gets twice as much horizontal space as `p2` and `p3`.

---

# Stacking and Placing Side by Side

`patchwork` provides two dedicated operators for the two most common layouts:

| Operator | Layout |
|----------|--------|
| `\|` | Place plots **beside** each other (same row) |
| `/` | **Stack** plots on top of each other (same column) |

## Side by side with `|`

```{r}
p1 | p2
```

## Stacked with `/`

```{r}
p1 / p2
```

---

# Nesting Layouts

Because `|` and `/` return patchwork objects themselves, they can be **nested** using parentheses to create complex multi-panel figures.

## One column beside a stacked pair

```{r}
p1 | (p2 / p3)
```

## A stacked pair beside another plot

```{r}
(p1 / p2) | p3
```

## Three-level nesting

```{r}
p1 | (p2 / (p3 | p4))
```

::: {.callout-tip}
**Reading nested patchwork expressions** — evaluate the innermost parentheses first, just like arithmetic. `p3 | p4` is a row, `p2 / (p3 | p4)` stacks `p2` above that row, and finally `p1 | ...` places `p1` to the left of the whole thing.
:::

---

# Annotating a Composition

A combined figure often needs a shared title, subtitle, caption, or tagged subplot labels. Use `plot_annotation()` for this.

## Shared title and caption

```{r}
(p1 | (p2 / p3)) +
  plot_annotation(
    title   = "A story about `mtcars`",
    subtitle = "Exploring displacement, horsepower, weight, and fuel economy",
    caption  = "Source: Motor Trend Car Road Tests (1974)"
  )
```

## Auto-tagging subplots

Patchwork can automatically label each panel so you can refer to them in text as (A), (B), etc.

```{r}
p1 + p2 + p3 +
  plot_annotation(tag_levels = "A")
```

Available tag styles:

| `tag_levels` | Labels |
|--------------|--------|
| `"A"` | A, B, C, ... |
| `"a"` | a, b, c, ... |
| `"1"` | 1, 2, 3, ... |
| `"I"` | I, II, III, ... |
| `"i"` | i, ii, iii, ... |

```{r}
p1 + p2 + p3 + p4 +
  plot_annotation(tag_levels = "I")
```

## Styling tags with `tag_prefix` and `tag_suffix`

```{r}
p1 + p2 + p3 + p4 +
  plot_annotation(
    tag_levels = "A",
    tag_prefix = "(",
    tag_suffix = ")"
  )
```

---

# Modifying Individual Panels in a Composition

After combining plots, you can still apply ggplot2 modifications to a specific panel using `[[i]]` indexing.

```{r}
combined <- p1 + p2 + p3

# Change the theme of the second panel only
combined[[2]] <- combined[[2]] + theme_minimal()

combined
```

---

# Collecting Legends

When multiple panels share the same aesthetic (e.g., the same color scale), their legends are redundant. Use `plot_layout(guides = "collect")` to merge them into one.

```{r}
# Two plots that both use fill = factor(cyl)
q1 <- ggplot(mtcars) +
  geom_point(aes(mpg, disp, color = factor(cyl))) +
  labs(title = "mpg vs. disp", color = "Cylinders")

q2 <- ggplot(mtcars) +
  geom_point(aes(hp, wt, color = factor(cyl))) +
  labs(title = "hp vs. wt", color = "Cylinders")

q1 + q2 + plot_layout(guides = "collect")
```

Without `guides = "collect"`, each panel carries its own legend — cluttering the figure. With it, the legend is deduplicated and placed at the right edge by default.

---

# Adding Empty Space with `plot_spacer()`

Sometimes you want a blank cell in your grid (e.g., to shift a plot to the right or leave room for a manual annotation). Use `plot_spacer()`:

```{r}
p1 + plot_spacer() + p2
```

```{r}
# Place p3 in the bottom-right of a 2x2 grid
plot_spacer() + plot_spacer() + plot_spacer() + p3
```

---

# Insets with `inset_element()`

`inset_element()` places one plot on top of another as an inset — like a zoomed-in detail or a small map overlay.

```{r}
p3 + inset_element(p1, left = 0.6, bottom = 0.6, right = 1, top = 1)
```

The `left`, `right`, `bottom`, and `top` arguments are proportions of the base plot's area (0 to 1). Here the inset occupies the upper-right 40% × 40% of `p3`.

---

# Putting It All Together — A Publication-Ready Figure

Combining all the tools we have learned: nested layouts, collected legends, auto-tags, and a shared annotation.

```{r}
#| fig-width: 10
#| fig-height: 8

final_figure <-
  ((q1 | q2) / (p2 | p4)) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title    = "Motor Trend Car Road Tests — `mtcars` Overview",
    subtitle = "Top row: continuous relationships by cylinder count | Bottom row: discrete summaries",
    caption  = "Data: Motor Trend (1974)",
    tag_levels = "A",
    tag_prefix = "(",
    tag_suffix = ")"
  ) &
  theme_minimal(base_size = 11)

final_figure
```

::: {.callout-note}
## The `&` operator
`& theme_minimal()` applies a theme to **all panels at once**, unlike `+` which only targets the last panel. Use `&` whenever you want a global style change across every plot in a patchwork.
:::

---

# Quick Reference

| Task | Code |
|------|------|
| Place side by side | `p1 \| p2` |
| Stack vertically | `p1 / p2` |
| Fill a 2×2 grid | `p1 + p2 + p3 + p4` |
| Control rows/cols | `+ plot_layout(nrow = 2)` |
| Relative widths | `+ plot_layout(widths = c(2, 1))` |
| Shared title/caption | `+ plot_annotation(title = "...")` |
| Auto-tag panels | `+ plot_annotation(tag_levels = "A")` |
| Collect legends | `+ plot_layout(guides = "collect")` |
| Blank cell | `plot_spacer()` |
| Inset plot | `+ inset_element(p, left, bottom, right, top)` |
| Style all panels | `& theme_minimal()` |
| Modify one panel | `combined[[2]] <- combined[[2]] + ...` |

---

# Further Reading

- [patchwork documentation](https://patchwork.data-imaginist.com/)
- [Plot Assembly guide](https://patchwork.data-imaginist.com/articles/guides/assembly.html) — `wrap_plots()`, `wrap_elements()`, nesting details
- [Controlling Layouts](https://patchwork.data-imaginist.com/articles/guides/layout.html) — `area()`, complex grid specifications
- [Annotation and Style](https://patchwork.data-imaginist.com/articles/guides/annotation.html) — per-level tagging, theme control
- [Multi-page Alignment](https://patchwork.data-imaginist.com/articles/guides/multipage.html) — aligning panels across `grid` and `cowplot` pages
