---
title: Lecture 4
subtitle: LaTeX Math Notation and Quarto Workflow for Capstone Papers
author:
  - name: Byeong-Hak Choe
    email: bchoe@geneseo.edu
    affiliations: SUNY Geneseo
date: 2026-05-04
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-overflow: wrap
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

# Learning goals

By the end of this lecture, students should be able to:

1. Write common mathematical expressions using LaTeX notation in Quarto.
2. Distinguish between inline math and display math.
3. Write regression equations clearly in an empirical paper.
4. Organize a capstone paper project using a reproducible Quarto workflow.
5. Understand why data files do **not** always need to be inside the website working directory.

# Part 1: LaTeX math notation in Quarto {background-color="#1c4982"}

## Why use LaTeX math notation?

In empirical research papers, mathematical notation helps us describe models, variables, hypotheses, and results clearly.

For example, instead of writing:

> Outcome equals intercept plus beta times treatment plus error.

We can write:

$$
Y_{it} = \beta_0 + \beta_1 X_{it} + \varepsilon_i
$$

This is shorter, cleaner, and more professional.

## Inline math

Use single dollar signs for math that appears inside a sentence.

```markdown
The coefficient $\beta_1$ measures the relationship between $X_{it}$ and $Y_{it}$.
```

This renders as:

The coefficient $\beta_1$ measures the relationship between $X_{it}$ and $Y_{it}$.

## Display math

Use double dollar signs for equations that should appear on their own line.

```markdown
$$
Y_{it} = \beta_0 + \beta_1 X_{it} + \varepsilon_i
$$
```

This renders as:

$$
Y_{it} = \beta_0 + \beta_1 X_{it} + \varepsilon_i
$$

## Greek letters

Many statistical and econometric models use Greek letters.

| LaTeX code | Output | Meaning |
|---|---:|---|
| `\alpha` | $\alpha$ | alpha |
| `\beta` | $\beta$ | beta |
| `\gamma` | $\gamma$ | gamma |
| `\delta` | $\delta$ | delta |
| `\varepsilon` | $\varepsilon$ | error term |
| `\sigma` | $\sigma$ | standard deviation |
| `\mu` | $\mu$ | mean |

Example:

```markdown
The error term is written as $\varepsilon_i$.
```

## Subscripts and superscripts

Use `_` for subscripts and `^` for superscripts.

```markdown
$Y_{it}$

$X_{it}$

$R^2$

$X_{it}^2$
```

These render as:

$Y_{it}$

$X_{it}$

$R^2$

$X_{it}^2$

Use curly braces `{}` when the subscript or superscript has more than one character.

Correct:

```markdown
$X_{it}$
```

Less clear:

```markdown
$X_{it}t$
```

## Fractions

Use `\frac{numerator}{denominator}`.

```markdown
$$
\bar{X} = \frac{1}{n}\sum_{i=1}^{n} X_{it}
$$
```

This renders as:

$$
\bar{X} = \frac{1}{n}\sum_{i=1}^{n} X_{it}
$$

## Summation notation

Use `\sum`.

```markdown
$$
\sum_{i=1}^{n} X_{it}
$$
```

This renders as:

$$
\sum_{i=1}^{n} X_{it}
$$

Example: sample mean

$$
\bar{X} = \frac{1}{n}\sum_{i=1}^{n} X_{it}
$$

## Square roots

Use `\sqrt{}`.

```markdown
$$
RMSE = \sqrt{\frac{1}{n}\sum_{i=1}^{n}(Y_{it} - \hat{y}_i)^2}
$$
```

This renders as:

$$
RMSE = \sqrt{\frac{1}{n}\sum_{i=1}^{n}(Y_{it} - \hat{y}_i)^2}
$$

## Hats, bars, and other accents

| LaTeX code | Output | Common meaning |
|---|---:|---|
| `\hat{Y}` | $\hat{Y}$ | predicted value |
| `\bar{X}` | $\bar{X}$ | sample mean |
| `\tilde{X}` | $\tilde{X}$ | transformed variable |
| `\widehat{\beta}` | $\widehat{\beta}$ | estimated coefficient |

Example:

```markdown
The fitted value is $\hat{Y}_i$.
```

## Parentheses and brackets

Common symbols:

```markdown
$(x)$

$[x]$

$\{x\}$

$\left( \frac{x}{y} \right)$
```

These render as:

$(x)$

$[x]$

$\{x\}$

$\left( \frac{x}{y} \right)$

The commands `\left` and `\right` automatically adjust the size of parentheses.

## Regression equation: basic linear model

A simple regression model can be written as:

$$
Y_{it} = \beta_0 + \beta_1 X_{it} + \varepsilon_i
$$

where:

- $Y_{it}$ is the outcome variable for observation $i$.
- $X_{it}$ is the explanatory variable.
- $\beta_0$ is the intercept.
- $\beta_1$ is the slope coefficient.
- $\varepsilon_i$ is the error term.

## Regression equation: multiple regression

A multiple regression model can be written as:

$$
Y_{it} = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \beta_3 X_{3i} + \varepsilon_i
$$

A more compact version is:

$$
Y_{it} = \beta_0 + \mathbf{X}_i'\boldsymbol{\beta} + \varepsilon_i
$$

where $\mathbf{X}_i$ is a vector of explanatory variables.

## Regression equation: panel data model

For panel data, we often use two subscripts:

- $i$ for unit, such as person, county, firm, or country.
- $t$ for time, such as year, month, or day.

Example:

$$
Y_{it} = \beta_0 + \beta_1 X_{it} + \gamma' Z_{it} + \alpha_i + \lambda_t + \varepsilon_{it}
$$

where:

- $Y_{it}$ is the outcome for unit $i$ at time $t$.
- $X_{it}$ is the main explanatory variable.
- $Z_{it}$ is a vector of control variables.
- $\alpha_i$ represents unit fixed effects.
- $\lambda_t$ represents time fixed effects.
- $\varepsilon_{it}$ is the error term.

## Log models

A log-linear model:

$$
\log(Y_{it}) = \beta_0 + \beta_1 X_{it} + \varepsilon_i
$$

A log-log model:

$$
\log(Y_{it}) = \beta_0 + \beta_1 \log(X_{it}) + \varepsilon_i
$$

In a log-log model, $\beta_1$ can often be interpreted as an elasticity:

> A 1% increase in $X$ is associated with approximately a $\beta_1$% change in $Y$.

## Hypotheses

Hypotheses can be written as:

$$
H_0: \beta_1 = 0
$$

$$
H_A: \beta_1 \neq 0
$$

Or, for a one-sided test:

$$
H_A: \beta_1 > 0
$$

## Common LaTeX mistakes

### Mistake 1: Forgetting dollar signs

Incorrect:

```markdown
The coefficient beta_1 is positive.
```

Correct:

```markdown
The coefficient $\beta_1$ is positive.
```

### Mistake 2: Using plain text inside math mode

Less ideal:

```markdown
$Y = beta0 + beta1 X$
```

Better:

```markdown
$Y = \beta_0 + \beta_1 X$
```

### Mistake 3: Missing curly braces

Less clear:

```markdown
$X_{it}t$
```

Better:

```markdown
$X_{it}$
```

# Part 2: Quarto workflow for a capstone paper {background-color="#1c4982"}

## Why use Quarto for a capstone paper?

Quarto allows you to combine:

- written explanation,
- code,
- tables,
- figures,
- mathematical notation,
- citations,
- and final output files,

all in one reproducible document.

A capstone paper should not only show results. It should also show a clear and reproducible workflow.

## Recommended project structure

A good capstone project folder may look like this:

```text
capstone-project/
в”‚
в”њв”Ђв”Ђ capstone-paper.qmd
в”‚
в”њв”Ђв”Ђ code/
в”‚   в”њв”Ђв”Ђ 01_data_cleaning.R
в”‚   в”њв”Ђв”Ђ 02_analysis.R
в”‚   в””в”Ђв”Ђ 03_modeling.R
в”‚
в”њв”Ђв”Ђ output/
в”‚   в”њв”Ђв”Ђ cleaned_data.csv
в”‚   в”њв”Ђв”Ђ regression_table.html
в”‚   в””в”Ђв”Ђ figures/
в”‚       в”њв”Ђв”Ђ figure_1.png
в”‚       в””в”Ђв”Ђ figure_2.png
в”‚
в””в”Ђв”Ђ website/
    в”њв”Ђв”Ђ index.qmd
    в””в”Ђв”Ђ _quarto.yml
```

This is only one possible structure. The most important idea is that your files should be organized intentionally.

## Important tip: data files can be outside the website directory

A common mistake is putting every file inside the website folder.

That is not always necessary.

For example, suppose your website folder is:

```text
capstone-project/website/
```

Your data file can still be outside that folder:

```text
capstone-project/data/my_data.csv
```

In your Quarto file inside the website folder, you can read the data using a relative path:

```r
data <- readr::read_csv("../data/my_data.csv")
```

The `..` means вЂњgo up one folder.вЂќ

So this path:

```text
../data/my_data.csv
```

means:

1. Start inside `website/`.
2. Go up to `capstone-project/`.
3. Go into `data/`.
4. Read `my_data.csv`.

## Why keep large data outside the website directory?

There are several reasons:

1. Large data files can make your website repository messy.
2. Some data files should not be publicly posted.
3. GitHub Pages is mainly for publishing webpages, not storing large raw datasets.
4. Keeping data separate makes your project easier to organize.
5. You can submit or share data separately through Brightspace, Google Drive, OneDrive, SharePoint, or Dropbox.

## Example: reading data from a folder outside the website

```{r}
# Example only
# This code assumes the data file exists at ../data/my_data.csv

# library(readr)
# my_data <- read_csv("../data/my_data.csv")
```

## Example: reading data from an online URL

If your data is public, you can also read it from a URL.

```{r}
# Example only
# library(readr)
# my_data <- read_csv("https://example.com/my_data.csv")
```

This can be useful when the data is publicly available and stable.

## Example: using an absolute path during development

During early development, you may use an absolute path on your computer:

```r
my_data <- readr::read_csv("/Users/yourname/Documents/capstone-data/my_data.csv")
```

However, this is not ideal for final submission because this path only works on your own computer.

A better final version is usually a relative path:

```r
my_data <- readr::read_csv("../data/my_data.csv")
```

Or a public/shareable URL:

```r
my_data <- readr::read_csv("https://example.com/my_data.csv")
```

## Working directory intuition

The working directory is the folder where R or Quarto looks for files by default.

In Quarto, the working directory is usually the folder that contains the `.qmd` file being rendered.

For example, if you render this file:

```text
capstone-project/website/index.qmd
```

then relative paths are interpreted from:

```text
capstone-project/website/
```

So:

```r
read_csv("data/my_data.csv")
```

looks for:

```text
capstone-project/website/data/my_data.csv
```

But:

```r
read_csv("../data/my_data.csv")
```

looks for:

```text
capstone-project/data/my_data.csv
```

## Recommended workflow for the capstone paper

## Step 1: Start with a clear research question

A weak research question:

> I want to study housing prices.

A stronger research question:

> How do neighborhood characteristics predict housing prices in Monroe County?

A still stronger version:

> How does school district quality relate to housing prices in Monroe County after controlling for property characteristics?

## Step 2: Organize the data

Recommended data workflow:

1. Start with raw data.
2. Clean and transform the data.
3. Save a cleaned version.
4. Use the cleaned version for visualization and modeling.

Example:

```text
raw data в†’ cleaning script в†’ cleaned data в†’ figures/tables/models
```

## Step 3: Write the paper structure

A typical empirical capstone paper may include:

```text
1. Introduction
2. Background and Motivation
3. Data
4. Methods
5. Results
6. Discussion
7. Conclusion
8. References
```

## Step 4: Write code in chunks

Example R code chunk:

````markdown
```{r}
library(tidyverse)

df <- read_csv("../data/my_data.csv")

df |> 
  summarise(
    avg_y = mean(y, na.rm = TRUE),
    avg_x = mean(x, na.rm = TRUE)
  )
```
````

## Step 5: Separate final results from scratch work

Your final capstone paper should include polished results.

It should not include every piece of trial-and-error code.

For example, do not include five failed versions of the same graph in the final paper. Keep only the final version and explain why it matters.

## Step 6: Render frequently

Do not wait until the deadline to render your Quarto file.

Render frequently while working.

This helps you catch problems early, such as:

- missing packages,
- broken file paths,
- syntax errors,
- missing images,
- incorrect YAML,
- or LaTeX math errors.

## Step 7: Check both the website and the submitted files

Before submitting, check:

1. Does the rendered HTML webpage open correctly?
2. Are all figures visible?
3. Are all tables visible?
4. Does the code run?
5. Are file paths correct?
6. Did you submit the `.qmd` file?
7. Did you submit required data files or links?
8. Did you include code files if your code is not fully embedded in the `.qmd` file?

# Part 3: Writing equations in the capstone paper {background-color="#1c4982"}

## Example methods section

Here is an example paragraph for a methods section:

> To examine the relationship between neighborhood characteristics and housing prices, I estimate the following regression model:

$$
\log(Price_i) = \beta_0 + \beta_1 SchoolQualitY_{it} + \beta_2 Size_i + \beta_3 Age_i + \varepsilon_i
$$

> The dependent variable is the log of housing price. The main explanatory variable is school quality. The model also controls for house size and age. The coefficient $\beta_1$ measures the association between school quality and housing prices, holding the other included variables constant.

## Example interpretation

Suppose the estimated model is:

$$
\log(Price_i) = 2.1 + 0.04 SchoolQualitY_{it} + 0.002 Size_i - 0.01 Age_i
$$

Then one possible interpretation is:

> Holding house size and age constant, a one-unit increase in school quality is associated with an approximately 4% increase in housing price.

## Example model with prediction

For a predictive modeling project, you may write:

$$
\hat{Y}_i = f(X_{it})
$$

where $f(\cdot)$ is a prediction function learned from the training data.

For example, in a random forest model:

$$
\hat{Y}_i = \frac{1}{B}\sum_{b=1}^{B} T_b(X_{it})
$$

where $T_b(X_{it})$ is the prediction from tree $b$, and $B$ is the number of trees.

# Part 4: Practical Quarto tips {background-color="#1c4982"}

## Tip 1: Use relative paths when possible

Good:

```r
read_csv("../data/my_data.csv")
```

Less ideal for final submission:

```r
read_csv("/Users/yourname/Desktop/my_data.csv")
```

## Tip 2: Do not upload private data to a public website

If your data contains sensitive or private information, do not publish it on GitHub Pages.

Instead, submit the data privately through Brightspace or share it through a secure cloud link.

## Tip 3: Keep file names simple

Good file names:

```text
capstone-paper.qmd
cleaned_data.csv
housing_model_results.csv
figure_1_price_distribution.png
```

Avoid file names like:

```text
final final real final paper!!!.qmd
my data (copy) version 7.csv
```

## Tip 4: Avoid spaces in file names

This is better:

```text
cleaned_data.csv
```

This can cause problems:

```text
cleaned data.csv
```

## Tip 5: Use comments in code

Good comments explain the purpose of the code.

```r
# Load cleaned data for the final analysis
df <- read_csv("../data/cleaned_data.csv")
```

Avoid comments that simply repeat the code.

```r
# Read csv
df <- read_csv("../data/cleaned_data.csv")
```

## Tip 6: Use section headings clearly

Use one `#` for major sections:

```markdown
# Introduction
```

Use two `##` for subsections:

```markdown
## Data source
```

Use three `###` for smaller subsections:

```markdown
### Variable construction
```

# Part 5: Mini checklist for students

Before submitting your capstone paper, ask:

- Did I clearly state my research question?
- Did I explain where my data came from?
- Did I describe important variables?
- Did I include meaningful tables and figures?
- Did I explain my model or method?
- Did I interpret the results in words?
- Did I discuss limitations?
- Did I include references?
- Did I render the Quarto file successfully?
- Did I submit the `.qmd` file, data files or links, and any separate code files?

# In-class practice

## Practice 1: Convert text to math notation

Write your model equation in LaTeX math notation:


## Practice 2: Fix the file path

Suppose your project is organized like this:

```text
capstone-project/
в”‚
в”њв”Ђв”Ђ data/
в”‚   в””в”Ђв”Ђ cleaned_data.csv
в”‚
в””в”Ђв”Ђ website/
    в””в”Ђв”Ђ index.qmd
```

You are writing code inside `index.qmd`.

What path should you use to read `cleaned_data.csv`?

Answer:

```r
readr::read_csv("../data/cleaned_data.csv")
```

## Practice 3: Write a model for your own project

Fill in the following model using variables from your own capstone project:

$$
Y_{it} = \beta_0 + \beta_1 X_{it} + \varepsilon_i
$$

Then explain:

- What is $Y_{it}$?
- What is $X_{it}$?
- What does $\beta_1$ mean?
- What is one limitation of this model?

# Closing summary

LaTeX math notation helps you write professional and precise research papers. Quarto helps you combine writing, code, figures, tables, and equations into one reproducible workflow.

For your capstone paper, focus on clarity:

- clear research question,
- clear data description,
- clear method,
- clear results,
- clear interpretation,
- and clear file organization.

A strong capstone paper is not just a collection of code and outputs. It is a complete research story supported by reproducible analysis.
